Towards atomistic resolution structure of phosphatidylcholine headgroup and glycerol backbone at different ambient conditions\footnote{Publication about results presented in the NMRlipids project}

### 2015-05-13 HGmodelMANUSCRIPT/HGmodel_ACStemplate.tex
Changes after fast read-through.

\subsubsection{Phospholipid bilayer with low hydration level}
The experimental order parameters available in the literature~\cite{dvinskikh05b,ulrich94,bechinger91} 
for the glycerol backbone and choline as a function of hydration level are shown in Fig.~\ref{ordPhydr}. 
The independently reported values for choline segments are in good agreement with each other (despite 
slight differences in temperature and acyl chain composition),
showing increase for both segments with decreasing hydration level. It should be noted that only 
absolute values were measured in the original experiments~\cite{dvinskikh05b,ulrich94,bechinger91}, but
we have included the signs measured in separate studies~\cite{hong95a,hong95b,gross97}. 
Consequently, the $\beta$ order parameter with negative sign actually increases with dehydration 
since the absolute value decreases~\cite{dvinskikh05b,ulrich94,bechinger91}.
Slight decrease for the glycerol backbone order parameters were observed with dehydration~\cite{dvinskikh05b}.
\begin{figure}[]
  \centering
  \includegraphics[width=7.3cm]{../DATAreportediINblog/dehydration_lp.pdf}
    \includegraphics[width=7.3cm]{../DATAreportediINblog/dehydration_smooth.pdf}
    \caption{\label{ordPhydr}
    The effect of dehydration on glycerol and choline order parameters in experiments.
    The magnitudes of order parameters are measured for DMPC ($^1$H-$^{13}$C NMR) at 314~K~\cite{dvinskikh05b}, 
    for POPC ($^2$H NMR) at 296~K~\cite{bechinger91} and for DOPC ($^2$H NMR) at 303~K~\cite{ulrich94}. 
    The signs are based on the measurements by Hong et al.~\cite{hong95a,hong95b} 
    and Gross et al.~\cite{gross97}.
    Note that to elucidate the relative change as a function of hydration level,
    the simulation results are shifted such that the (smaller) $S_\mathrm{CH}$
    matches (within $\pm$0.01) the experimental value at full hydration;
    the shift magnitudes for each of the force fields are listed ($S_\mathrm{CH}$+shift) in the y-label.
  }
  	\todo{Markus prepared two versions of this plot. The one on left has all the data points joint by lines;
	the one on right shows smooth trend lines (cubic polynomials) to ease qualitative comparison.
	Samuli likes the one on left, Markus the one on right. Do others have opinions?}
\end{figure}

Lipid bilayer dehydration has been studied also with molecular dynamics simulations~\cite{mashl01,pertsin05,pertsin07,eun09,eun10,schneck12},
typically motivated by the  discussion about the origin of the ``hydration repulsion''~\cite{israelachvili,israelachvili96,sparr11}.
However, the used simulation models are not typically compared to the experimental choline and glycerol backbone
order parameters (except by Mashl et al.~\cite{mashl01}).
In Fig.~\ref{ordPhydr} the glycerol backbone and choline order parameters are shown as a function of hydration level for the CHARMM36, 
MacRog and GAFFlipid models (having the most realistic atomistic resolution structures) together with the Berger based model 
(which is the most used lipid model). 
To elucidate changes the simulation results are shifted to match better with experiments with full hydration.
The magnitude of shifting is written in the y-axis label.
The choline order parameter increase with dehydration is seen in all
models despite of some additional fluctuations. Thus the choline order parameter response to dehydration can be
intepreted to be in qualitative agreement with experiments.
The situation is significantly more complicated for the glycerol backbone segments 
and none of the models can be interpreted to qualitatively agree with experiments
for all the segmets.

The qualitative agreement with experiments in all simulation models for the $\alpha$ and $\beta$ order parameters  
as a function of hydration indicates that the structural response of the choline headgroup to dehydration is somewhat realistic
despite the unrealistic structures at full hydration. 
The most likely explanation is that the choline group
orients more parallel to the membrane plane with dehydration due to restricted interlamellar space. 
Indeed, the P--N (phoshate phosphorus to choline nitrogen) vector angle with membrane normal shows an increase for
all models as a function of dehydration (Fig.~\ref{PNangle}). However,
the amount of increase depends on the model. Especially the DLPC simulations with Berger model
predict significantly stronger P--N vector tilt compared with the other models. The Berger model
has also generally larger P--N vector angles and its choline order parameters are more off from
experiments than other models. Thus the relatively modest tilting with dehydration
predicted by MacRog, CHARMM36 and GAFFlipid is probably more realistic.

@@

  \caption{\label{PNangle}
    The angle between membrane normal and P--N vector as function of
    hydration level calculated from different simulations.
  }
\end{figure}

>>>>

\subsubsection{Phospholipid bilayer with low hydration level}
The experimental order parameters available in the literature~\cite{dvinskikh05b,ulrich94,bechinger91} 
for the glycerol backbone and choline as a function of hydration level are shown in Fig.~\ref{ordPhydr}. 
The independently reported values for the choline segments are in good agreement with each other (despite 
slight differences in temperature and acyl chain composition),
and both ($\beta$ and $\alpha$) show increase with decreasing hydration level. It should be noted that only 
absolute values were measured in the original experiments~\cite{dvinskikh05b,ulrich94,bechinger91}, but
we have included the signs measured in separate studies~\cite{hong95a,hong95b,gross97}. 
Consequently, the $\beta$ order parameter with negative sign actually increases with dehydration 
since the absolute value decreases~\cite{dvinskikh05b,ulrich94,bechinger91}.
Slight decrease for the glycerol backbone order parameters have been observed with dehydration~\cite{dvinskikh05b}.
\begin{figure}[]
  \centering
  \includegraphics[width=7.3cm]{../DATAreportediINblog/dehydration_lp.pdf}
    \includegraphics[width=7.3cm]{../DATAreportediINblog/dehydration_smooth.pdf}
    \caption{\label{ordPhydr}
    The effect of dehydration on glycerol and choline order parameters in experiments.
    The magnitudes of order parameters are measured for DMPC ($^1$H-$^{13}$C NMR) at 314~K~\cite{dvinskikh05b}, 
    for POPC ($^2$H NMR) at 296~K~\cite{bechinger91} and for DOPC ($^2$H NMR) at 303~K~\cite{ulrich94}. 
    The signs are based on the measurements by Hong et al.~\cite{hong95a,hong95b} 
    and Gross et al.~\cite{gross97}.
    Note that to elucidate the relative change as a function of hydration level,
    the simulation results are shifted such that the (smaller) $S_\mathrm{CH}$
    matches (within $\pm$0.01) the experimental value at full hydration;
    the shift magnitudes for each of the force fields are listed ($S_\mathrm{CH}$+shift) in the y-label.
  }
  	\todo{Markus prepared two versions of this plot. The one on left has all the data points joint by lines;
	the one on right shows smooth trend lines (cubic polynomials) to ease qualitative comparison.
	Samuli likes the one on left, Markus the one on right. Do others have opinions?}
\end{figure}

Lipid bilayer dehydration has been studied also with molecular dynamics simulations~\cite{mashl01,pertsin05,pertsin07,eun09,eun10,schneck12},
typically motivated by the  discussion about the origin of the ``hydration repulsion''~\cite{israelachvili,israelachvili96,sparr11}.
However, the used simulation models are typically not compared to the experimental choline and glycerol backbone
order parameters (except by Mashl et al.~\cite{mashl01}).
In Fig.~\ref{ordPhydr} the glycerol backbone and choline order parameters are shown as a function of hydration level for the CHARMM36, 
MacRog and GAFFlipid models (having the most realistic atomistic resolution structures) together with the Berger based model 
(which is the most used lipid model). 
To elucidate changes the simulation results are shifted to match better with experiments with full hydration.
The magnitude of shifting is written in the y-axis label.
The increase of the choline order parameters with dehydration is seen in all
models despite of some additional fluctuations. Thus the choline order parameter response to dehydration can be
intepreted to be in qualitative agreement with experiments.
The situation is significantly more complicated for the glycerol backbone segments 
and none of the models can be interpreted to qualitatively agree with experiments
for all the segmets.

The qualitative agreement with experiments in all simulation models for the $\alpha$ and $\beta$ order parameters  
as a function of hydration indicates that the structural response of the choline headgroup to dehydration is somewhat realistic
despite the unrealistic structures at full hydration. 
The most likely explanation is that the choline group
orients more parallel to the membrane plane with dehydration due to restricted interlamellar space. 
Indeed, the P--N (phoshate phosphorus to choline nitrogen) vector angle with membrane normal shows an increase for
all models as a function of dehydration (Fig.~\ref{PNangle}). However,
the amount of increase depends on the model. Especially the DLPC simulations with Berger model
predict significantly stronger P--N vector tilt compared with the other models. The Berger model
has also generally larger P--N vector angles and its choline order parameters are more off from
experiments than the other three models. Thus the relatively modest tilting with dehydration
predicted by MacRog, CHARMM36 and GAFFlipid is probably more realistic.

@@

  \caption{\label{PNangle}
    The average angle between membrane normal and P--N vector as function of
    hydration level calculated from different simulations.
  }
\end{figure}
